\bfseries A Variant of the Sandwich theorem for two-sided limits

### 2025-07-01 lecturenotes_aima203/ch6_aima203.tex
ch6 and article

\section{Numerical Differentiation}
The general method for deriving the numerical differentiation formulae is to differentiate the interpolating polynomial. Hence, corresponding to each of the Interpolating formula derived, we may derive a formula for the derivative.

\subsection{Newton's forward difference formulae}
The Newton's forward difference formula is:
\begin{equation}
  \label{newtforward6}
 \begin{gathered}
  y_n(x) = y_0 + u \Delta y_0 + \frac{u(u-1)}{2!}\, \Delta ^2 y_0 + \frac{u(u-1)(u-2)}{3!}\, \Delta ^3 y_0 + ... \\[1mm]
  + \frac{u(u-1)(u-2)...(u-(n-1))}{n!}\, \Delta ^n y_0
\end{gathered}
\end{equation}
where, $x=x_0 + uh$ and $h = x_{i+1} - x_{i}$. Differentiating ~\ref{newtforward6} with respect to $x$,

@@

\subsection{Newton's backward difference formula}
In a similar way, different formulae can be derived by starting with other interpolation formulae. Thus, Newton's backward difference formula gives
\begin{equation}
  \label{newtbackwarddif}
  \left [ \frac{dy}{dx} \right ] _{x=x_n} = \frac{1}{h} \left [\nabla \, y_n + \frac{1}{2} \nabla ^2 \, y_n + \frac{1}{3} \nabla ^3 \, y_n + ... \right ]
\end{equation}
and
\begin{equation}
  \label{newtbackwarddif2}
  \left [ \frac{d^2y}{dx^2} \right ] _{x=x_n} = \frac{1}{h^2} \left [\nabla^2 \, y_n + \nabla ^3 \, y_n + \frac{11}{12} \nabla ^4 \, y_n + \frac{5}{6} \nabla ^5 \, y_n + ... \right ]
\end{equation}
\begin{remark}
   If the $x$-values of the data points are equally spaced then it is common to use Newton's interpolation.
\end{remark}
\subsection{Exercise}
Find the $dy/dx$ and $d^2y/dx^2$ at $x=1.2, \; x=1.6, \; x=2.2$ from the data: \\[1mm]
$(1, 2.7183), \; (1.2, 3.3201), \; (1.4, 4.0522), \; (1.6, 4.9530), \; (1.8, 6.0496), \; (2.0, 7.3891), \; (2.2, 9.0250)$.
\section{Numerical Integration}
The general problem of numerical integration may be stated as: Given a set of data points of a function $y=f(x)$, where $f(x)$ is not known explicitly, it is required compute the value of the definite integral $$I = \int_a^b \; y\,dx$$ As with the case of numerical differentiation, one replaces $f(x)$ by an interpolating polynomial $\phi(x)$ and obtains, on integration, an approximate value of the definite integral. Thus, different integration formulae can be obtained depending upon the type of the interpolation formula used. Approximating $y$ by Newton's forward difference formula, we obtained
$$ I = \int_{x_0}^{x_n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, dx$$
Since $x=x_0 +ph$, $dx = hdp$. When $x=x_0$, $p=0$ and $x=x_n$, $p=n$. Then,
$$ I = \int_{0}^{n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, h\,dp$$
Integrating we get,
\begin{equation}
  \label{numint}
  \begin{aligned}[b]
    I &= h \left [ py_0 + \frac{p^2}{2}\, \Delta y_0 + \left ( \frac{p^3}{3} - \frac{p^2}{2} \right ) \, \frac{\Delta^2 y_0}{2} +  \left ( \frac{p^4}{4} - p^3 + p^2 \right ) \frac{\Delta ^3 \, y_0}{6} + ... \right ]_0^n \\[1mm]
    &= nh \left [ y_0 + \frac{n}{2}\, \Delta y_0 + \left ( \frac{2n^2-3n}{12} \right ) \, \Delta^2\, y_0 +  \left ( \frac{n^3 - 4n^2 +4n}{24} \right ) \, \Delta ^3 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}

@@

\item Evaluate using Simpson's $3/8$ rule: $\displaystyle \int_0^1 \, \frac{dx}{1+x}$ with $h=1/6$.

\item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$. Then obtain a better estimate using \textbf{Romberg's method}.

\item Compute $\displaystyle I\left (\frac{\pi}{4} \right ), \; I\left (\frac{\pi}{8} \right ), \; I\left (\frac{\pi}{8}, \frac{\pi}{4} \right )$ for : $I = \displaystyle \int_0^{\pi/4} \, cos^2x\, dx$.

\item Using Trapezoidal rule and Simpson's $1/3$ rule evaluate the double integral $$\int_{_2}^2 \, \int_0^4 \; (x^2 -xy + y^2) \; dx\, dy.$$

\end{enumerate}
\subsection{Boole's and Weddle's Rules, \textit{Extra material}}
To incorporate the fourth order difference as well we need $5$ data points $x_0, ..., x_4$ and we need to integrate from $x_0$ to $x_4$. Then we obtain Boole's formula:
\begin{equation}
  \label{Boole's}
  \displaystyle \int_{x_0}^{x_4} \; y\,dx = \frac{2h}{45}\, (7y_0 + 32y_1 + 12y_2+32y_3+7y_4)
\end{equation}
If we incorporate differences u to six order then we obtain Weddle's formula as follows:
\begin{equation}
  \label{Weddle's}
  \displaystyle \int_{x_0}^{x_6} \; y\,dx = \frac{3h}{10}\, (y_0 + 5y_1 + y_2+6y_3+y_4+5y_5+y_6)
\end{equation}

@@

\begin{figure}[h]
  \centering
\includegraphics[width=12cm, height=9cm]{gaussian.png}
\end{figure}
\subsection{Exercise}
\begin{enumerate}
\item Derive the Gauss quadrature formula for $n=2$ and apply it to evaluate the integral $\displaystyle \int_{-1}^1 \, \frac{dx}{1+x^2}$.

\item Use the three point Gauss formula to evaluate the integral $\displaystyle \int_0^1 \, \frac{dx}{1+x}$.
\end{enumerate}
\section{Double Integration}
Formulae for the evaluation of double integral can be obtained by repeatedly applying the trapezoidal and Simpson's rules derived earlier.
\subsection{Trapezoidal Rule}
\begin{equation*}
  I = \int_{y_j}^{y_{j+1}}\; \int_{x_i}^{x_{i+1}} \; f(x,y)\;dx\,dy
\end{equation*}
where, $x_{i+1} = x_i + h$ and $y_{j+1} = y_j + k$.
By repeated application of trapezoidal rule to we get,
\begin{equation}
  \label{trapdouble}
  \begin{aligned}
    I &= \frac{h}{2} \,\int_{y_j}^{y_{j+1}}\; [f(x_i,y) + f(x_{i+1},y) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f(x_i,y_j) + f(x_{i+1},y_j) + f(x_i,y_{j+1}) + f(x_{i+1},y_{j+1}) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f_{i,j} + f_{i+1, j} + f_{i,j+1} + f_{i+1,j+1} ]\;dx\,dy
  \end{aligned}
\end{equation}

\subsection{Simpson's Rule}
\begin{equation*}
  I = \int_{y_{j-1}}^{y_{j+1}}\; \int_{x_{i-1}}^{x_{i+1}} \; f(x,y)\;dx\,dy
\end{equation*}
where, $x_{i+1} = x_i + h$ and $y_{j+1} = y_j + k$.
By repeated application of Simpson's rule to we get,
\begin{equation}
  \label{simpdouble}
  \begin{aligned}
    I &= \frac{h}{3} \,\int_{y_{j-1}}^{y_{j+1}}\; [f(x_{i-1},y) + 4f(x_i,y) + f(x_{i+1},y) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{9} \, [f(x_{i-1},y_{j-1}) + 4f(x_{i-1},y_j) + f(x_{i-1},y_{j+1}) \\[1mm]
    & 4\{f(x_i,y_{j-1}) + 4f(x_i,y_j) + f(x_i,y_{j+1})\} + f(x_{i+1},y_{j-1}) + 4f(x_{i+1},y_j) + f(x_{i+1},y_{j+1})]
  \end{aligned}
\end{equation}
\vspace{1cm}

\begin{figure}[h]
  \centering
\includegraphics[width=15cm, height=11cm]{doubleint.png}
\end{figure}

\end{document}

>>>>

\section{Numerical Differentiation}
The general method for deriving the numerical differentiation formulae is to \textbf{differentiate the interpolating polynomial}. Hence, corresponding to each of the Interpolating formula derived, we may derive a formula for the derivative.

\subsection{Newton's forward difference formulae}
The Newton's forward difference formula is:
\begin{equation}
  \label{newtforward6}
 \begin{gathered}
  y_n(x) = y_0 + u \Delta y_0 + \frac{u(u-1)}{2!}\, \Delta ^2 y_0 + \frac{u(u-1)(u-2)}{3!}\, \Delta ^3 y_0 + ... \\[1mm]
  + \frac{u(u-1)(u-2)...(u-(n-1))}{n!}\, \Delta ^n y_0
\end{gathered}
\end{equation}
where, $\mathbf{x=x_0 + uh}$ and $h = x_{i+1} - x_{i}$. Differentiating ~\ref{newtforward6} with respect to $x$,

@@

\subsection{Newton's backward difference formula}
In a similar way, different formulae can be derived by starting with other interpolation formulae. Thus, Newton's backward difference formula gives: where $\mathbf{x = x_n + uh}$
\begin{equation}
  \label{newtbackwarddif}
  \left [ \frac{dy}{dx} \right ] _{x=x_n} = \frac{1}{h} \left [\nabla \, y_n + \frac{1}{2} \nabla ^2 \, y_n + \frac{1}{3} \nabla ^3 \, y_n + ... \right ]
\end{equation}
and
\begin{equation}
  \label{newtbackwarddif2}
  \left [ \frac{d^2y}{dx^2} \right ] _{x=x_n} = \frac{1}{h^2} \left [\nabla^2 \, y_n + \nabla ^3 \, y_n + \frac{11}{12} \nabla ^4 \, y_n + \frac{5}{6} \nabla ^5 \, y_n + ... \right ]
\end{equation}
\begin{remark}
   If the $x$-values of the data points are equally spaced then it is common to use Newton's interpolation.
\end{remark}
\subsection{Exercise}
\begin{enumerate}
\item Find the $dy/dx$ and $d^2y/dx^2$ at $x=1.2, \; x=1.6, \; x=2.2$ from the data:
  \begin{table}[h]
    \centering
    \begin{tabular}{|c|c|c|c|c|c|}
      \hline
      x& 1& 1.2& 1.4& 1.6&1.8\\
      y &2.7183& 3.3201& 4.0522& 4.9530& 6.60496\\
      \hline
    \end{tabular}
  \end{table}

\item The distance traversed by a particle at different times are given below. Find the velocity and the acceleration of the particle at $t=0.1$ seconds.
  \begin{table}[h]
    \centering
    \begin{tabular}{|c|c|c|c|c|}
      \hline
      t& 0.0& 0.1& 0.2& 0.3\\
      x & 3.01 & 3.16& 3.29& 3.36\\
      \hline
    \end{tabular}
  \end{table}

\item From the given data points $(x,y)$, find $\displaystyle \frac{dy}{dx}$ at $x=2$ using the natural cubic spline method.
$(2,11), \; (3,49), \; (4,123)$.
\end{enumerate}

\section{Numerical Integration}
The general problem of numerical integration may be stated as: Given a set of data points of a function $y=f(x)$, where $f(x)$ is not known explicitly, it is required compute the value of the definite integral $$I = \int_a^b \; y\,dx$$ As with the case of numerical differentiation, one replaces $f(x)$ by an interpolating polynomial $\phi(x)$ and obtains, on integration, an approximate value of the definite integral. Thus, different integration formulae can be obtained depending upon the type of the interpolation formula used. Approximating $y$ by Newton's forward difference formula, we obtained
$$ I = \int_{x_0}^{x_n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, dx$$
Since $x=x_0 +ph$, $dx = hdp$. When $x=x_0$, $p=0$ and $x=x_n$, $p=n$. Then,
$$ I = \int_{0}^{n} \; \left [ y_0 + p \Delta y_0 + \frac{p(p-1)}{2}\, \Delta^2 y_0 + \frac{p(p-1)(p-2)}{6}\, \Delta ^3 y_0 + ... \right ] \, h\,dp$$
Integrating we get,
\begin{equation}
  \label{numint}
  \begin{aligned}[b]
    I &= h \left [ py_0 + \frac{p^2}{2}\, \Delta y_0 + \left ( \frac{p^3}{3} - \frac{p^2}{2} \right ) \, \frac{\Delta^2 y_0}{2} +  \left ( \frac{p^4}{4} - p^3 + p^2 \right ) \frac{\Delta ^3 \, y_0}{6} + ... \right ]_0^n \\[1mm]
    &= nh \left [ y_0 + \frac{n}{2}\, \Delta y_0 + \left ( \frac{2n^2-3n}{12} \right ) \, \Delta^2\, y_0 +  \left ( \frac{n^3 - 4n^2 +4n}{24} \right ) \, \Delta ^3 \, y_0 + ... \right ]
  \end{aligned}
\end{equation}

@@

\item Evaluate using Simpson's $3/8$ rule: $\displaystyle \int_0^1 \, \frac{dx}{1+x}$ with $h=1/6$.
\end{enumerate}
\subsection{Boole's and Weddle's Rules, \textit{Extra material}}
To incorporate the fourth order difference as well we need $5$ data points $x_0, ..., x_4$ and we need to integrate from $x_0$ to $x_4$. Then we obtain Boole's formula:
\begin{equation}
  \label{Boole's}
  \displaystyle \int_{x_0}^{x_4} \; y\,dx = \frac{2h}{45}\, (7y_0 + 32y_1 + 12y_2+32y_3+7y_4)
\end{equation}
If we incorporate differences u to six order then we obtain Weddle's formula as follows:
\begin{equation}
  \label{Weddle's}
  \displaystyle \int_{x_0}^{x_6} \; y\,dx = \frac{3h}{10}\, (y_0 + 5y_1 + y_2+6y_3+y_4+5y_5+y_6)
\end{equation}

@@

\begin{figure}[h]
  \centering
\includegraphics[width=12cm, height=9cm]{gaussian.png}
\end{figure}
\subsection{Exercise}
\begin{enumerate}
 \item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$. Then obtain a better estimate using \textbf{Romberg's method}.

\item Compute $\displaystyle I\left (\frac{\pi}{4} \right ), \; I\left (\frac{\pi}{8} \right ), \; I\left (\frac{\pi}{8}, \frac{\pi}{4} \right )$ for : $I = \displaystyle \int_0^{\pi/4} \, cos^2x\, dx$.
\item Derive the Gauss quadrature formula for $n=2$ and apply it to evaluate the integral $\displaystyle \int_{-1}^1 \, \frac{dx}{1+x^2}$.

\item Use the three point Gauss formula to evaluate the integral $\displaystyle \int_0^1 \, \frac{dx}{1+x}$. Compare this result with that obtained by Simpson's 1/3-rule with $h=0.125$.
\end{enumerate}
\section{Double Integration}
Formulae for the evaluation of double integral can be obtained by repeatedly applying the trapezoidal and Simpson's rules derived earlier.
\subsection{Trapezoidal Rule}
\begin{equation*}
  I = \int_{y_j}^{y_{j+1}}\; \int_{x_i}^{x_{i+1}} \; f(x,y)\;dx\,dy
\end{equation*}
where, $x_{i+1} = x_i + h$ and $y_{j+1} = y_j + k$.
By repeated application of trapezoidal rule to we get,
\begin{equation}
  \label{trapdouble}
  \begin{aligned}
    I &= \frac{h}{2} \,\int_{y_j}^{y_{j+1}}\; [f(x_i,y) + f(x_{i+1},y) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f(x_i,y_j) + f(x_{i+1},y_j) + f(x_i,y_{j+1}) + f(x_{i+1},y_{j+1}) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{4} \, [f_{i,j} + f_{i+1, j} + f_{i,j+1} + f_{i+1,j+1} ]\;dx\,dy
  \end{aligned}
\end{equation}

\subsection{Simpson's Rule}
\begin{equation*}
  I = \int_{y_{j-1}}^{y_{j+1}}\; \int_{x_{i-1}}^{x_{i+1}} \; f(x,y)\;dx\,dy
\end{equation*}
where, $x_{i+1} = x_i + h$ and $y_{j+1} = y_j + k$.
By repeated application of Simpson's rule to we get,
\begin{equation}
  \label{simpdouble}
  \begin{aligned}
    I &= \frac{h}{3} \,\int_{y_{j-1}}^{y_{j+1}}\; [f(x_{i-1},y) + 4f(x_i,y) + f(x_{i+1},y) ]\;dx\,dy \\[1mm]
      &= \frac{hk}{9} \, [f(x_{i-1},y_{j-1}) + 4f(x_{i-1},y_j) + f(x_{i-1},y_{j+1}) \\[1mm]
    & 4\{f(x_i,y_{j-1}) + 4f(x_i,y_j) + f(x_i,y_{j+1})\} + f(x_{i+1},y_{j-1}) + 4f(x_{i+1},y_j) + f(x_{i+1},y_{j+1})]
  \end{aligned}
\end{equation}
\begin{figure}[h]
  \centering
\includegraphics[width=15cm, height=11cm]{doubleint.png}
\end{figure}

\textbf{Questions:} Using Trapezoidal rule and Simpson's $1/3$ rule evaluate the double integral $$\int_{_2}^2 \, \int_0^4 \; (x^2 -xy + y^2) \; dx\, dy.$$
\newpage

\begin{figure}[h]
  \centering
\includegraphics[width=15cm, height=9cm]{trapezoidexample.png}
\end{figure}

\begin{figure}[h]
  \centering
\includegraphics[width=15cm, height=10cm]{simpsonexample.png}
\end{figure}
\end{document}

### 2025-07-01 sandwich_theorem/extended_sandwich_theorem.tex
ch6 and article

\documentclass[a4paper,twoside,12pt]{article}
\linespread{1}                     % For line spacing
\usepackage{amssymb}                 % For AMS symbol
\usepackage{amsmath,amsthm,latexsym}                 % This is useful for the matrix $\begin{bmatrix} a & b\\ 0 & c\end{bmatrix}$
\usepackage{verbatim}                % For comments in paragraph
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{epsfig}              % For imbeding graph
\usepackage{color}                  % Color for the text
\usepackage[normal]{caption2}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{multicol}
\setlength{\columnsep}{0.8cm}
\usepackage{makeidx}
\usepackage[normal]{subfigure}
\usepackage{url}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{plain}
%%%%%%%%%%%%%
\textheight23cm%25.0cm
\textwidth16.3cm
\oddsidemargin-0.1cm
\evensidemargin0.1cm
%\topmargin-1cm
\headsep.3cm
%\usepackage{pstricks}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}% Clear header/footer
\fancyhead[OC]{The Extended Sandwich Theorem; S. Thakuri}%Author on Odd page, Centred
\fancyhead[EC]{The Extended Sandwich Theorem; S. Thakuri }% Title on Even page, Centred
\fancyfoot[C]{\thepage}%

@@

\title{\bfseries The Extension of the Sandwich theorem for limits}

@@

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\begin{document}
\linenumbers
\vspace{2mm}
{\Large
\begin{center}
\bf{\LARGE \bfseries A Variant of the Sandwich theorem for limits}
\end{center}}
\begin{center}
Sandesh Thakuri$^{1}$, Bishnu Hari Subedi$^{2}$
\end{center}

\begin{center}
{\footnotesize
  $^{1}$Department of Artificial Intelligence, SoE, Kathmandu University, Nepal \\[1mm]
 \(^{2}\)Central Department of Mathematics, IoST, Tribhuvan University, Kirtipur, Nepal \\[2mm]
 Correspondence to: Sandesh Thakuri, Email: sandesh.775509@cdmath.tu.edu.np \\[1mm]
\textit{ 2020 Mathematics Subject Classification. 26A03}
}
\end{center}

\vspace{5mm}
\noindent
\textbf{Abstract:} {The scope of the Sandwich theorem for limit in analysis is limited to two sided-limits only. This paper extends the scope of the Sandwich theorem by considering one-sided limit as well and gives a more general version of  the Sandwich theorem by making the criterion of the Sandwich theorem more loose, i.e it relaxes the condition of the Sandwich theorem.}

\section{Introduction}
The Sandwich theorem is simple yet powerful tool in analysis to determine and to analyze the limit of a function at a given point. We can leverage the known limits to calculate the unknown limits. Suppose, we know the limits of \(g(x)\) and \(h(x)\) at \(x=c\) to be the same limit \(L\) and here \(f(x)\) happens to be sandwich between \(g(x)\) and \(h(x)\) in some neighborhood of \(c\). Then we can conclude the limit of \(f(x)\) at \(x=c\) as \(L\) by the Sandwich theorem which is as follows.

@@

As an illustration, we know the limit of \(1/x\) is \(0\) as \(x
\to \infty\). This implies the limit of \(-1/x\) is also \(0\) as \(x \to \infty\). These are the known limits.\\
The value of sine function lies between \(-1\) and \(1\) so that \[\displaystyle -\frac{1}{x} \leq \frac{sinx}{x} \leq \frac{1}{x}\]
Now, \(\displaystyle \lim_{x \to \infty} -\frac{1}{x}=0\) and \(\displaystyle \lim_{x \to \infty} \frac{1}{x}=0\) so we can use the Sandwich theorem to conclude that \(\displaystyle \lim_{x \to \infty} \frac{sinx}{x}=0\).
\vspace{5mm}
\section{Limitation of the Sandwich Theorem}
The sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation of the sandwich theorem.
We know \(\displaystyle - \left | x \right | \leq sinx \leq \left | x \right | \) which implies \[-\frac{\left | x \right |}{x} \leq \frac{sinx}{x} \leq \frac{\left | x \right |}{x}\]

But we cannot conclude the limit of \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) using the Sandwich Theorem with the bounds above. It is so, because the \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist. \\[2mm]
The \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist because left hand limit and right hand limit of \(\displaystyle \frac{|x|}{x}\) at x=0 are different. They are different but they are finite.\\
\(\displaystyle \lim_{x \to 0^-} \frac{|x|}{x}=-1\) \hspace{5mm} and \hspace{5cm} \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\).

\vspace{5mm}
\section{The Thakuri's Variant of Sandwich Theorem}
\begin{theorem}[The Thakuri's Extended Sandwich Theorem]
Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself. Suppose also that $$\lim_{x \to c^-} g(x) = \lim_{x \to c^+} h(x) =L  .$$ Then $\displaystyle \lim_{x \to c} f(x)=L$.
\end{theorem}

Now using this Extended Sandwich theorem it can be conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
 \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
 \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[3mm]

@@

 \section{Remarks on the our variant of Sandwich theorem}
 \begin{itemize}
 \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$, the purpose of our sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem. This limit can be determined without the sandwich theorem as well.

@@

\vspace{5mm}
 \section{Proof of the Extended Sandwich Theorem}

@@

\item Rudin variant for one-sided limit is analogous to the classical Sandwich theorem for two-side limits. Our variant is not analogous to the classical Sandwich theorem. Rudin's variant has Sandwich theorem for left-hand side limit and for right-hand side limit separately. Our variant has left-hand limit and right-had limit only to loosen the criterion and it is not for the one-sided limit. It is for the two-sided limit.
  \end{itemize}

\vspace{5mm}
\bibliographystyle{aomalpha}
\begin{thebibliography}{9}
    \bibitem{thomas}
Thomas, G. B. (2017) \emph{Thomas' Calculus}. Pearson India Education Services Pvt. Ltd, India.
\end{thebibliography}

>>>>

\documentclass[a4paper,twoside,12pt]{article}
\linespread{1}                     % For line spacing
\usepackage{amssymb}                 % For AMS symbol
\usepackage{amsmath,amsthm,latexsym}                 % This is useful for the matrix $\begin{bmatrix} a & b\\ 0 & c\end{bmatrix}$
\usepackage{verbatim}                % For comments in paragraph
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{epsfig}              % For imbeding graph
\usepackage{color}                  % Color for the text
\usepackage[normal]{caption2}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{multicol}
\setlength{\columnsep}{0.8cm}
\usepackage{makeidx}
\usepackage[normal]{subfigure}
\usepackage{url}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{plain}
%%%%%%%%%%%%%
\textheight23cm%25.0cm
\textwidth16.3cm
\oddsidemargin-0.1cm
\evensidemargin0.1cm
%\topmargin-1cm
\headsep.3cm
%\usepackage{pstricks}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}% Clear header/footer
\fancyhead[OC]{The Thakuri's Variant of Sandwich Theorem; S. Thakuri}%Author on Odd page, Centred
\fancyhead[EC]{The Thakuri's Variant of Sandwich Theorem ; S. Thakuri }% Title on Even page, Centred
\fancyfoot[C]{\thepage}%

@@

\title{\bfseries A Variant of the Sandwich theorem for two-sided limits}

@@

%---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
\begin{document}
\linenumbers
\vspace{2mm}
{\Large
\begin{center}
\bf{\LARGE \bfseries A Variant of the Sandwich theorem for two-sided limits}
\end{center}}
\begin{center}
Sandesh Thakuri$^{1}$, Bishnu Hari Subedi$^{2}$
\end{center}

\begin{center}
{\footnotesize
  $^{1}$Department of Artificial Intelligence, SoE, Kathmandu University, Nepal \\[1mm]
 \(^{2}\)Central Department of Mathematics, IoST, Tribhuvan University, Kirtipur, Nepal \\[2mm]
 Correspondence to: Sandesh Thakuri, Email: sandesh.775509@cdmath.tu.edu.np \\[1mm]
\textit{2020 Mathematics Subject Classification. 26A03}
}
\end{center}

\vspace{5mm}
\noindent
\textbf{Abstract:} {The criterion for the classical Sandwich theorem for two-sided limits is that the two-side limits must exists for the bounding functions. We show that this criterion can be relaxed. We show that it is sufficient for the existence of left-hand limit for lower bound function and the existence of right-hand limit for the upper bound function; and of-course they must be equal. This paper relaxes the criterion of the classical Sandwich theorem by replacing the two-sided limits with one-sided limits in the criterion and gives a new variant of the Sandwich theorem.  While Rudin has formulated the Sandwich theorem for the one-sided limits, it still doesn't relax the criterion of the Sandwich theorem for two-sided limits. He has incorporated the one-sided limits for the Sandwich theorem for one-sided limits but has not relax the condition for the Sandwich theorem for two-sided limits.}
\section{Introduction}
The Sandwich theorem is simple yet powerful tool in analysis to determine and to analyze the limit of a function at a given point. We can leverage the known limits to calculate the unknown limits. Suppose, we know the limits of \(g(x)\) and \(h(x)\) at \(x=c\) to be the same limit \(L\) and here \(f(x)\) happens to be sandwich between \(g(x)\) and \(h(x)\) in some neighborhood of \(c\). Then we can conclude the limit of \(f(x)\) at \(x=c\) as \(L\) by the Sandwich theorem for two-sided limits. The Sandwich theorem for two-sided limits is simply called the Sandwich theorem which is as follows.

@@

As an illustration, we know the limit of \(1/x\) is \(0\) as \(x
\to \infty\). This implies the limit of \(-1/x\) is also \(0\) as \(x \to \infty\). These are the known limits.\\
The value of sine function lies between \(-1\) and \(1\) so that \[\displaystyle -\frac{1}{x} \leq \frac{sinx}{x} \leq \frac{1}{x}\]
Now, \(\displaystyle \lim_{x \to \infty} -\frac{1}{x}=0\) and \(\displaystyle \lim_{x \to \infty} \frac{1}{x}=0\) so we can use the Sandwich theorem to conclude that \(\displaystyle \lim_{x \to \infty} \frac{sinx}{x}=0\).
\vspace{5mm}
\section{Limitation of the Sandwich Theorem}
The Sandwich theorem requires the existence of the two sided limits for the lower bound function $g(x)$ and the upper bound function $h(x)$. This is the limitation of the sandwich theorem. \\
We know \(\displaystyle - \left | x \right | \leq sinx \leq \left | x \right | \) which implies \[-\frac{\left | x \right |}{x} \leq \frac{sinx}{x} \leq \frac{\left | x \right |}{x}\]
But we cannot conclude the limit of \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) using the Sandwich Theorem with the bounds above. It is so, because the \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist. \\[2mm]
The \(\displaystyle \lim_{x \to 0} \frac{|x|}{x}\) does not exist because left hand limit and right hand limit of \(\displaystyle \frac{|x|}{x}\) at x=0 are different. They are different but they are finite.\\
\(\displaystyle \lim_{x \to 0^-} \frac{|x|}{x}=-1\) \hspace{5mm} and \hspace{5cm} \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\). \\[3mm]
We are not claiming that the limitation of the Sandwich theorem is not being able to determine the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\). We are stating that the limitation of Sandwich theorem is its requirement of  the existence of the two sided limits for the bounding functions thus not being able to determine the above limit with the above bounds.
\vspace{5mm}
\section{The Thakuri's Variant of Sandwich Theorem}
\begin{theorem}[The Thakuri's Variant Sandwich Theorem]
Suppose $g(x) \leq f(x) \leq h(x)$ in some open interval containing $c$, expect possibly at $x=c$ itself. Suppose also that $$\lim_{x \to c^-} g(x) = \lim_{x \to c^+} h(x) =L  .$$ Then $\displaystyle \lim_{x \to c} f(x)=L$.
\end{theorem}

Now using this variant Sandwich theorem we can use the above bounds to conclude that \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}=1\) as follows. \\[2mm]
 \(\displaystyle \lim_{x \to 0^-} -\frac{|x|}{x}=1\) \hspace{7cm}
 \(\displaystyle \lim_{x \to 0^+} \frac{|x|}{x}=1\) \\[3mm]

@@

 \section{Remarks on the our variant of Sandwich theorem}
 \begin{itemize}
 \item  While the \(\displaystyle \lim_{x \to 0} \frac{sinx}{x}\) can be determined by the sandwich theorem using the tighter-bound $cosx \leq sinx/x \leq 1$, the purpose of our variant of sandwich theorem is not to assert that this limit cannot be determined by the Sandwich theorem. This limit can be determined without the sandwich theorem as well.

@@

\vspace{5mm}
 \section{Proof of the Variant of the Sandwich Theorem}

@@

\item Rudin variant for one-sided limit is analogous to the classical Sandwich theorem for two-side limits. Our variant is not analogous to the classical Sandwich theorem. Rudin's variant has Sandwich theorem for left-hand side limit and for right-hand side limit separately. Our variant has left-hand limit and right-had limit only to loosen the criterion and it is not for the one-sided limit. It is for the two-sided limit.
  \end{itemize}
\vspace{5mm}
\bibliographystyle{aomalpha}
\begin{thebibliography}{9}
\bibitem{hardy}
  Hardy, G. H. (1961) \emph{A Course of Pure Mathematics, 10th edition}. Cambridge University Press, London.
\bibitem{rudin}
  Rudin, W. (1976) \emph{Principles of Mathematical Analysis, 3rd edition}. McGraw Hill, London.
\bibitem{tao}
  Tao, T. (2023) \emph{Analysis I, 2nd edition}. Springer Nature, Singapore.
\bibitem{thomas}
  Thomas, G. B. (2017) \emph{Thomas' Calculus}. Pearson India Education Services Pvt. Ltd, India.
\end{thebibliography}
